Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-02-21 paper_jasa.tex
pdf version of lineup for experiment 1 and 2 added

\begin{figure}[htp]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk1_300_10_12_1.png}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  \blue {Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 15 which was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. For this lineup, 15 out of 16 observers correctly pick the data plot. }

@@

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. \blue{The actual data plot location is 5. None of the 65 observers could detect the actual plot.}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk2_100_25_5_3.png}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:continuous}.)}
       \label{fig:test_continuous}
\end{figure}

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? \blue{The actual data plot location is 1. Out of 31 observers 13 observers could detect the actual plot.}

@@

\subsection{Model fitting}
 For each parameter combination, the effect is computed as

\[
E=\sqrt {n} \cdot \beta/\sigma.
\]
\noindent
The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept, i.e. 
 $W_{\ell i} = E_{\ell i}$. Instead of fitting an intercept, we make use of a fixed  offset of $\log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different skill levels of subjects are accounted for  by allowing subject-specific random  slopes for  effect ($E$).

@@

Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice:

\begin{enumerate} \itemsep 0in
\item Most different plot
\item Visible trend
\item Clustering visible
\item Other
\end{enumerate}

\noindent with the possibility to use more than one. The task requested subjects to identify the plot that had the largest slope, which would correspond to choosing ``visible trend'' (2) as the reason for their choice. Reasons 1 or 3 would be indicative of Type III error. Figure \ref{fig:choice_reason} explores the reasons subjects gave for their choices. If there were no Type III errors committed, we would expect that people overwhelmingly using ``visible trend'' as their reason, or at least, when they use this reason they overwhelmingly correctly choose the actual data plot. This is not what we see. At left, are the reasons subjects gave for their choices --- 123 means that they gave all three reasons. The horizontal axis shows proportion of times that subjects correctly chose the actual data plot, and the reasons are sorted from most accurate to least accurate. The size of the point corresponds to the number of subjects putting this as the reason. Subjects that chose all three reasons almost always chose the actual data plot. This was followed by using 1 and 3, and then 1 and 4. The most common reasons given were reasons 1-3 individually, and the accuracy for these reasons ranged from 75\% for reason ``most different plot'' to 60\% for ``visible trend''.
At right is a simplified view, containing just the four possible reasons -- if the subject chose one of these, regardless if they also chose another reason it is counted. ``Visible trend'' comes in third. This is strong evidence that for many subjects even though they are correctly choosing the data plot, often they are cueing to other structure in the plot than the trend, making a Type III error.

>>>>

\begin{figure}[htp]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk1_300_10_12_1.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  \blue {Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 15 which was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. For this lineup, 15 out of 16 observers picked the actual data plot. }

@@

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. \blue{The actual data plot location is 5. None of the 65 observers picked the actual plot while 46 observers picked plot 18 which has the lowest $p$-value among all the plots in this lineup as we can see in Figure \ref{fig:P-val_log2} for Experiment 2.}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk2_100_25_5_3.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:continuous}.)}
       \label{fig:test_continuous}
\end{figure}

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? \blue{The actual data plot location is 1. Out of 31 observers 13 observers picked the actual plot.}

@@

\subsection{Model fitting}
 For each parameter combination, the effect is computed as 
\[
E=\sqrt {n} \cdot \beta/\sigma.
\]
\noindent
The model in Equation \ref{eqn:mixed} is fit using $E$ as the only fixed effect covariate without intercept, i.e. 
 $W_{\ell i} = E_{\ell i}$. Instead of fitting an intercept, we make use of a fixed  offset of $\log(0.05/0.95)$ so that the estimated power has a fixed lower limit at 0.05 (Type-I error) when $E=0$. Different skill levels of subjects are accounted for  by allowing subject-specific random  slopes for  effect ($E$).

@@

%Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice: 
%
%
%\begin{enumerate} \itemsep 0in
%\item Most different plot
%\item Visible trend
%\item Clustering visible
%\item Other
%\end{enumerate}
%
%
%\noindent with the possibility to use more than one.

\blue{Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice: (1) Most different plot (2) Visible trend (3) Clustering visible (4) Other, with the option to use more than one.} The task requested subjects to identify the plot that had the largest slope, which would correspond to choosing ``visible trend'' (2) as the reason for their choice. Reasons 1 or 3 would be indicative of Type III error. Figure \ref{fig:choice_reason} explores the reasons subjects gave for their choices. If there were no Type III errors committed, we would expect that people overwhelmingly using ``visible trend'' as their reason, or at least, when they use this reason they overwhelmingly correctly choose the actual data plot. This is not what we see. At left, are the reasons subjects gave for their choices --- 123 means that they gave all three reasons. The horizontal axis shows proportion of times that subjects correctly chose the actual data plot, and the reasons are sorted from most accurate to least accurate. The size of the point corresponds to the number of subjects putting this as the reason. Subjects that chose all three reasons almost always chose the actual data plot. This was followed by using 1 and 3, and then 1 and 4. The most common reasons given were reasons 1-3 individually, and the accuracy for these reasons ranged from 75\% for reason ``most different plot'' to 60\% for ``visible trend''.
At right is a simplified view, containing just the four possible reasons -- if the subject chose one of these, regardless if they also chose another reason it is counted. ``Visible trend'' comes in third. This is strong evidence that for many subjects even though they are correctly choosing the data plot, often they are cueing to other structure in the plot than the trend, making a Type III error.
